Delivering Reliable Software in the European Scientific Arena: Evolution of Methodologies and New Trend Analysis

### 2018-02-08 paper-ieee-trans.tex
minor rephrases

@@

The software quality policies \cite{indigo-d31} cover: 1) the identification
and description of the \emph{SQA requirements} that the produced software need 
to comply with, and 2) the \emph{quality metrics}, to monitor each software product's 
behaviour throughout the development, release and production stages.

@@

The SQA requirements cover the set of conventions and recommendations that pave the way for
an adequate development, timely deliver and reliable operation of the software components
produced. The SQA criteria is in continuous evolution, reviewed periodically to match the
project's needs.

\begin{itemize}
\item \textit{Code style}.
Code style standards are enforced for every software component in the project stack. Community
most-adopted or \textit{de-facto standard} is initially recommended, however the developer teams eventually
choose the set of guidelines to comply with. The ultimate goal for code style assessment is to
improve the readability and reusability of the source code produced under the scope of the
project. Figure \ref{fig:fig_codestyle} showcases the code style standards and their popularity
over the software components contributing to the project.

\item \textit{Unit and Functional testing}.
Changes involving the addition of new functionalities are requested to be tested. \textit{Regression
testing} in this context is done enforcing the successful execution of the previous
functionality tests already present in the code, so to uncover any malfunction introduced by
the changes. On the other hand, \textit{unit testing} completes the source code testing evaluation by
focusing on the code's internal design. Project's SQA criteria sets a recommended threshold of
70\% coverage for unit testing. In Figure \ref{fig:fig_unittest}, unit testing coverage is compared
for each software component over the two major INDIGO--DataCloud releases, showing an increment in the 
averaged value of {\sl INDIGO--2} release. As it can be seen, a very small set of products with low coverage
values highly impacted the overall result, although a significant part of these products are upstream
contributions or products with codebases not fully under the control of the project. Nevertheless,
60\% of the products were over this threshold while 80\% of them had values above
the 50\% coverage.

\item \textit{Integration testing}. Software components usually interact with other services while on
operation state. Integration testing deals with the interactions among coupled software components or
parts of a system that cooperates to achieve a given functionality. This type of testing might be 
complex and, based on the project's experience, tough to be tackled automatically. The aim is to 
guarantee the overall operation of the component, focused on the services it intefaces with, whenever
new functionalities are introduced.

\item \textit{Code review}.
Represents the last step in the change management pipeline, once the candidate change has 
successfully passed through the testing methods described previously. It implies the human-based
revision of the proposed change to discuss its adequance in terms of scope, objective fulfillment, 
documentation completeness, etc. On approval, the candidate change is definitely merged into
production.

\item \textit{Documentation}.
Last but not least, the SQA criteria sets the path for the adoption of the developed software by
defining the documentation content, according to the target audience. These requirements also 
promote the automation, both in terms of documentation creation and service deployment. The
\textit{documentation is treated as code}, using a markup language, automatically rendered and
uploaded to online repositories \cite{indigo-gitbook}. Thus documentation is portable and 
human--reviewed, using the same workflow as the code does, validating the changes before being
updated into the production repository.

@@

\textit{Maintenance and user support metrics} are key for continuously improving the response
to issues reported by external users. Feeback is collected from outside helpdesks, such as 
EGI's GGUS \cite{ggus}, and the INDIGO--DataCloud's GitHub Issues tracker.

@@

\begin{figure}[ht]
\centering
\includegraphics[width=0.5\textwidth, height=50mm]{images/jenkins_CI_builds.png}
\caption{Evolution of the number of builds triggered as part of the CI implementation.}
\label{fig:fig_jenkins_CI_builds}
\end{figure}

\subsubsection{Continuous Integration}
\label{subsec:ci}
The INDIGO--DataCloud project promoted the application of a CI scenario that enforces,
for any piece of software produced, the testing requirements defined in the SQA criteria,
as described in Section \ref{subsec:sqa}. Such environment requires an automation
ready--to--go infrastructure where the different technologies involved -- source code
management platform, automation server and containerization -- interact with each
other to trigger the source code validation pipeline for each candidate change. The
pipeline is mainly comprised by the code style validation, unit and functional testing
coverage. The CI pipeline is complemented with additional quality checks, such as 
integration tests for the software components where automation of this type of testing
is applicable, or security linters for the static analysis of suspicious
constructs that could lead to security risks. Metrics gathering also takes part of
the CI pipeline so to have a per-change trend evolution of the common source code 
related metrics, such as the source lines of code (SLOC) or the cyclomatic complexity.

@@

Figure \ref{fig:fig_jenkins_CI_builds} shows the evolution, throughout the 
INDIGO--DataCloud's first and second releases, of the required test types builds since the
implementation of the CI infrastructure. Automated functional testing coverage were not
available for all the software stack, thus the associated number of builds are fewer. The
executions are not precisely aligned as checks can be manually triggered, through Jenkins
interface, useful for instance whenever a failed check needs to be rebuilt.

@@

Two applications from "Definition of Support to Research Communities" workpackage, 
DisVis \cite{disvis} and PowerFit \cite{powerfit}, were
integrated into a similar CI/CD pipeline described in section \ref{sec:devops}.
Figure \ref{fig:fig_disvis} shows the pipeline for the Disvis application.

@@

The released software is tested in production environments through the
Staged--Rollout process, before inclusion in the EGI's CloudManagement Distribution (CMD). 
Selected resource providers are requested to install
the most updated stable versions of the software, and to give access to them to their users. The
Staged--Rollout process is key to detect and mitigate issues that could only
appear in production environments.

>>>>

\begin{figure*}[ht]
\centering
\includegraphics[width=\textwidth]{images/unittest.png}
\caption{Comparison of the unit testing coverage values for the {\sl INDIGO--DataCloud} software stack over the {\sl INDIGO--1} and {\sl INDIGO--2} major releases.}
\label{fig:fig_unittest}
\end{figure*}

@@

The software quality policies were initially defined in the first deliverable document of the
project \cite{indigo-d31} and periodically reviewed to be endowed with applicable insights of 
software engineering practices, responses to evolving needs emanated from the user communities and 
feedback obtained from development teams. These policies cover the 1) identification
and description of the \emph{SQA requirements} that the produced software need 
to comply with, and the 2) \emph{quality metrics}, to monitor each software product's 
behaviour throughout the development, release and post-release stages.

@@

The SQA criteria is a collection of conventions and recommendations that pave the way for
the adequate development, timely deliver and reliable operation of the software components
produced. It emphasizes the quality requirements and best practices to be applied at the
development phase, such as style and testing compliance or human code reviews, to protect the
production source code versions. Besides, the criteria promotes the adoption of the software 
by providing minimum requirements of documentation content and stressing the usage of automated
solutions for software deployment.

\begin{itemize}
\item \textit{Code style}.
The ultimate goal for code style assessment is to improve the readability and reusability of the source code produced under the scope of the
project. Code style standards are enforced for every software component in the project stack. Community
most-adopted or \textit{de-facto standards} are initially recommended, however the development teams eventually
choose the set of guidelines to comply with.  Figure \ref{fig:fig_codestyle} highlights the code style standards and their popularity
over the INDIGO--DataCloud software stack.

\item \textit{Unit and Functional testing}.
Changes involving the addition of new functionalities are requested to be tested. \textit{Regression
testing} in this context is done enforcing the successful execution of the previous
functionality tests already present in the code, so to uncover any malfunction introduced by
the changes. On the other hand, \textit{unit testing} completes the source code testing evaluation by
focusing on the code's internal design. The SQA criteria sets a recommended threshold of
70\% coverage for unit testing. In Figure \ref{fig:fig_unittest}, unit testing coverage is compared
for each software component over the two major INDIGO--DataCloud releases, denoting an incremental 
trend over the course of the project. As depicted in the Figure, 50\% of the components were over the
70\% threshold while slightly less than 80\% of the product stack had values over 50\% coverage.

\item \textit{Integration testing}. Software components usually interact with other services while on
operation state. Integration testing deals with the interactions among coupled software components or
parts of a system that cooperates to achieve a given functionality. This type of testing might be 
complex and, based on the project's experience, tough to be tackled automatically. The aim is to 
guarantee the overall operation of the component with regards to the services it interfaces with,
whenever new functionalities are introduced.

\item \textit{Code review}.
Represents the last step in the change management pipeline, once the candidate change has 
successfully passed through the testing methods described previously. It implies the human-based
revision of the proposed change to discuss its adequance in terms of scope, objective fulfillment, 
documentation completeness, etc. On approval, the candidate change is definitely merged into
production. Human code reviewing is paramount in the software quality assessment, specially 
important when the SQA testing requirements are fully automated. Secure code reviews are
done at this stage, assessing common vulnerabilities from inputs coming from automated linters
and manual dynamic application security testing.

\item \textit{Documentation}.
The SQA criteria sets the path for the adoption of the developed software by
defining the documentation content, according to the target audience. These requirements also 
promote the automation, both in terms of documentation creation and service deployment. The
\textit{documentation is treated as code}, using a markup language, automatically rendered and
uploaded to online repositories \cite{indigo-gitbook}. Thus documentation is portable and 
human--reviewed, using the same workflow as the code does, validating the changes before being
updated into the production repository.

@@

\textit{Maintenance and user support metrics} are key for continuously improving the response
to issues reported by external users. Feeback is collected from outside helpdesks, such as 
EGI's GGUS \cite{ggus}, and the GitHub Issues tracker.

@@

\begin{figure}[ht]
\centering
\includegraphics[width=0.5\textwidth, height=50mm]{images/jenkins_CI_builds.png}
\caption{Evolution of the total number of testing builds triggered automatically as part of the Jenkins CI implementation.}
\label{fig:fig_jenkins_CI_builds}
\end{figure}

\subsubsection{Continuous Integration (CI)}
\label{subsec:ci}
The INDIGO--DataCloud project promoted the application of a CI scenario that enforces,
for any piece of software produced, the testing requirements defined in the SQA criteria,
as described in Section \ref{subsec:sqa}. Such environment requires an automation
ready--to--go infrastructure where the different services and technologies involved interact with each
other to trigger the source code validation pipeline for each candidate change. The
pipeline is mainly comprised by the code style validation, unit and functional testing
coverage. The CI pipeline is complemented with additional quality checks, such as 
integration tests for the software components where automation of this type of testing
is applicable, or security linters for the code static analysis of suspicious
constructs that could lead to security risks. Metrics gathering also takes part of
the CI pipeline so to have a per-change trend evolution of the common source code 
related metrics, such as the source lines of code (SLOC) or the cyclomatic complexity.

@@

Figure \ref{fig:fig_jenkins_CI_builds} shows the evolution, throughout the 
INDIGO--DataCloud's first and second releases, of the required test types builds since the
implementation of the CI infrastructure. Automated functional testing coverage were not
available for all the software stack, thus the associated number of builds are fewer.

@@

Two applications from "Definition of Support to Research Communities" work package, 
DisVis \cite{disvis} and PowerFit \cite{powerfit}, were
integrated into a similar CI/CD pipeline described in section \ref{sec:devops}.
Figure \ref{fig:fig_disvis} shows the pipeline for the Disvis application.

@@

The released software is tested in production environments through the
Staged--Rollout process, before inclusion in the EGI's Cloud Middleware Distribution (CMD). 
Selected resource providers are requested to install
the most updated stable versions of the software, and to give access to them to their users. The
Staged--Rollout process is key to detect and mitigate issues that could only
appear in production environments.
